---
title: "Day 2: Introduction to Raster Data"
teaching: 0
exercises: 0
questions:
- "What is the coordinate reference system for the maximum temperature raster? Is it the same as the coordinate reference system for the minimum temperature raster? Visually explore the patterns in minimum and maximum monthly temperature differs across Wisconsin"
objectives:
- "Importing raster data; checking coordinate reference system"
keypoints:
- "First key point. Brief Answer to questions. (FIXME)"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
solutions <- FALSE
```

##Load in and explore the maximum and minimum temperature rasters for Wisconsin.

```{r, echo=solutions, include=solutions}
library(raster)
library(ggplot2)
library(gdal)
```



```{r, echo=solutions, include=solutions}
mintemp <- raster("../data/mintemp_monthcold_wi.tif")
summary(mintemp)
```


```{r, echo=solutions, include=solutions}
maxtemp <- raster("../data/maxtemp_monthwarm_wi.tif")
summary(maxtemp)
```


##Are the coordinate reference systems the same in these two rasters?

```{r, echo=solutions, include=solutions}
mintemp
maxtemp

#or

crs(mintemp)
crs(maxtemp)
```

Yes, the CRS for both rasters is the same. Both are in longlat projection, datum WGS84.

##Create visualizations of both minimum and maximum temperature rasters using `ggplot2`

```{r, echo=solutions, include=solutions}
mintemp_df <- as.data.frame(mintemp, xy=TRUE)
maxtemp_df <- as.data.frame(maxtemp, xy=TRUE)

ggplot() +
    geom_raster(data = mintemp_df , aes(x = x, y = y, fill = mintemp_monthcold_wi)) +
    scale_fill_viridis_c() +
    coord_quickmap()
ggsave("03-day2-mintempraster.png", path="../fig/")

ggplot() +
    geom_raster(data = maxtemp_df , aes(x = x, y = y, fill = maxtemp_monthwarm_wi)) +
    scale_fill_viridis_c() +
    coord_quickmap()
ggsave("03-day2-maxtempraster.png", path="../fig/")

```

###Bonus

You can add the following aesthetics to a `ggplot` to customize the backgrounds, axis labels, title, legend label, and move the legend position to the bottom of the plot:


```{r, eval=FALSE}
ggplot() +
    geom_*() +
    labs(title = "Your Title", x="X Label", y="Y Label", fill="Legend Label") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")

```

Explore what these aesthetics do with your plot of maximum monthly temperature.

```{r, echo=solutions, include=solutions}
ggplot() +
    geom_raster(data = maxtemp_df , aes(x = x, y = y, fill = maxtemp_monthwarm_wi)) +
    scale_fill_viridis_c() +
    coord_quickmap() +
    labs(title = "Maximum Monthly Temperature: WI", x="Longitude", y="Latitude", fill="Max. Temp (C)") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
ggsave("03-day2-maxtempraster-sol.png", path="../fig/")

```



## Describe what patterns you observe across the state for coldest and warmest monthly temperatures in Wisconsin. What do you hypothesize is the effect of Lake Michigan?
